Design and Implementation of a Multithreaded Virtual Machine for Executing Linear Logic Programs

### 2014-07-25 abstract.tex
Update.

We have developed a virtual machine for Linear Meld, a forward-chaining linear logic programming language
where programs can be executed in parallel. In Linear Meld, a program is seen as a database of logical facts and a set of derivation rules.
The database of facts is partitioned by the nodes of a graph structure which leads
to parallelism when nodes are executed simultaneously.
Due to the foundations on linear logic, rules can retract facts in a declarative and structured fashion, leading
to more expressive programs.
We present the design and implementation of the virtual machine that executes on multicores,
with a focus on thread management, code organization, fact indexing, rule execution, and database organization for efficient fact insertion,
lookup and deletion. Our results show that the virtual machine is not only
capable of scaling programs with up to 16 threads but also exhibits interesting performance results when
compared against other programming languages.

>>>>

\begin{abstract}
Linear Meld is a concurrent forward-chaining linear logic programming
language where logical facts can be asserted and retracted in a
structured way. In Linear Meld, a program is seen as a database of
logical facts and a set of derivation rules. The database of facts is
partitioned by the nodes of a graph structure which leads to
parallelism when nodes are executed simultaneously. Due to the
foundations on linear logic, rules can retract facts in a declarative
and structured fashion, leading to more expressive programs. We
present the design and implementation of the virtual machine that we
implemented to run Linear Meld on multicores, with particular focus on
thread management, code organization, fact indexing, rule execution,
and database organization for efficient fact insertion, lookup and
deletion. Our results show that the virtual machine is
capable of scaling programs with up to 16 threads and also exhibits
interesting scalar performance results due to our indexing optimizations.
\end{abstract}

\category{D.3.4}{PROCESSORS}{Interpreters}
\category{D.3.4}{PROCESSORS}{Run-time environments}
\category{D.1.3}{PROGRAMMING TECHNIQUES}{Concurrent Programming}[Parallel Programming]

\terms{Design, Languages, Performance}

\keywords{Linear Logic, Virtual Machine, Implementation}

### 2014-07-25 conclusions.tex
Update.

We have presented a parallel virtual machine for executing forward-chaining linear logic programs, with particular focus
on parallel scheduling on multicores, rule execution and database organization for fast insertion,
lookup, and deletion of linear facts.
Our experimental results show that the VM is able to scale the execution of programs when run with up to 16 threads.
Although this is still a work in progress, the VM also fairs well against other programming languages.
Moreover, since LM programs are concurrent by default, we can easily get better performance from the start by executing
them with multiple threads.

In the future, we intend to take advantage of linear logic to perform whole-program optimizations,
including computing program invariants, loop detection in rules and bypass of rule priorities. For instance,
we want to detect during compile time which might be the best data structure for each predicate depending in how it
is used in the rules.

>>>>

\section{Conclusions}

We have presented a parallel virtual machine for executing
forward-chaining linear logic programs, with particular focus on
thread management, code organization, fact indexing, rule execution
and database organization for fast insertion, lookup, and deletion of
linear facts. Experimental results show that our VM is able to scale
the execution of programs when run with up to 16 threads.
Our results also show the importance of having an efficient indexing mechanisms
for facts. With our dynamic indexing, the VM automatically detects which predicates
need to be indexed in order to improve performance.
Due to these and other optimizations, the VM fairs relatively well against other programming languages, including
compiled languages. Moreover, since LM programs are
concurrent by default, we can easily get better performance from the
start by executing them with multiple threads.
As further work, we want to improve parallel scalability and
take advantage of linear logic to perform whole-program
optimizations, including computing program invariants, loop detection
in rules and bypass of rule priorities.

We think that our virtual machine is an excellent starting point to make
logic programming more desirable in the data-mining, machine learning and
distributed/parallel programming community. Moreover, our virtual machine can
be easily extended to execute over computer networks or to execute programs on really big datasets.
In a nutshell, LM provides a concise way to describe
graph-based algorithms that can be more easily reasoned about, a clear advantage over
competing systems.
%% For instance, we want to detect during compile time which might be the
%% best data structure for each predicate depending in how it is used in
%% the rules.

### 2014-07-25 language.tex
update figs.
Update.

Linear Meld (LM) is a logic programming language that offers a declarative and structured way to manage state.
A program consists of a database of facts and a set of derivation rules. The database includes persistent
and linear facts. Persistent facts cannot be deleted, while linear facts can be asserted and retracted.

The dynamic (or operational) semantics of LM are identical to Datalog.
Initially, we populate the database with the program's axiom facts and then determine which derivation rules can be applied by using the current database. Once a rule is applied, we derive new facts, which are then added to the database.
If a rule uses linear facts, they are retracted from the database.
The program stops when \emph{quiescence} is achieved, that is, when rules no longer apply.

Each fact is a predicate on a tuple of \emph{values}, where the type of the predicate prescribes the types of the arguments.
LM rules are type-checked using the predicate declarations in the header of the program. LM has a simple type system that includes types such as
\emph{node}, \emph{int}, \emph{float}, \emph{string}, \emph{bool}. Recursive types such as \emph{list X} and \emph{pair X; Y} are
also allowed.
Each rule in LM has a defined priority that is inferred from its position in the source file.
Rules at the beginning of the file have higher priority. We consider all
the new facts that have been not considered yet to create a set of \emph{candidate rules}.
The set of candidate rules is then applied (by priority) and updated as new facts are derived.

@@

\begin{figure}[t]
{\footnotesize
\begin{Verbatim}[numbers=right]
type left(node, node).
type right(node, node).
type linear value(node, int, int).
type linear replace(node, int, int).

replace(A, K, New),
value(A, K, Old)
   -o value(A, K, New). // we found our key

replace(A, RKey, RValue),
value(A, Key, Value),
!left(A, B),
RKey < Key // go left
   -o value(A, Key, Value),
      replace(B, RKey, RValue).

replace(A, RKey, RValue),
value(A, Key, Value),
!right(A, B),
RKey > Key // go right
   -o value(A, Key, Value),
      replace(B, RKey, RValue).

// binary tree configuration
value(@3, 3, 3). value(@1, 1, 1). value(@0, 0, 0).
value(@2, 2, 2). value(@5, 5, 5). value(@4, 4, 4).
value(@6, 6, 6).
!left(@1, @0). !left(@3, @1). !left(@5, @4). 
!right(@1, @2). !right(@3, @5). !right(@5, @6).

replace(@3, 6, 7). // replace value of key 6 to 7
\end{Verbatim}
}
\caption{Binary tree dictionary: replacing a key's value.}
  \label{code:btree_replace}
\end{figure}

We now present an example LM program in Fig.~\ref{code:btree_replace} that implements the key update operation for a binary tree
represented as a key/value dictionary.
We first declare all the predicates (lines 1-4), which represent the kinds of facts we are going to use.
Predicate \texttt{left/2} and \texttt{right/2} are persistent while \texttt{value/3} and \texttt{replace/3} are linear.
The \texttt{value/3} predicate assigns a key/value pair to a binary tree node and the \texttt{replace/3} predicate
represents an update operation that updates the key in the second argument to the value in the third argument.

The algorithm uses three rules for the three cases of updating a key's value: the first rule performs the update
(lines 6-8); the second rule recursively picks the left branch for the update operation (lines 10-15); and the third
rule picks the right branch (lines 17-22).
The axioms of this program are presented in lines 24-29 and they describe the initial binary tree configuration,
including keys and values. By having the \texttt{update(@3, 6, 7)} axiom instantiated at the root node $@3$, we intend to
change the value of key 6 to 7.
Note that when writing rules or axioms, persistent facts are preceded with a \texttt{!}.

Figure~\ref{fig:btree_trace} represents the trace of the algorithm. Note that the program database is partitioned
by the tree nodes using the first argument of each fact. In Fig.~\ref{fig:btree_trace}a we present the database
filled with the program's axioms. Next, we follow the right branch using rule 3 since $6 > 3$ (Fig.~\ref{fig:btree_trace}b).
We then use the same rule again in Fig.~\ref{fig:btree_trace}c where we finally reach the key 6. Here, we apply rule 1 and
\texttt{value(@6, 6, 6)} is updated to \texttt{value(@6, 6, 7)}.

\begin{figure*}[]
        \centering
        \begin{subfigure}[b]{0.45\textwidth}
                \includegraphics[width=\textwidth]{btree_trace1}
                \caption{Initial database. Replace axiom instantiated at root node $@3$.}
                \label{fig:btree_trace1}
        \end{subfigure}%
        ~
        \begin{subfigure}[b]{0.45\textwidth}
                \includegraphics[width=\textwidth]{btree_trace2}
                \caption{After applying rule 3 at node $@3$. Replace fact sent to node $@5$.}
                \label{fig:btree_trace2}
        \end{subfigure}\\
        \begin{subfigure}[b]{0.45\textwidth}
                \includegraphics[width=\textwidth]{btree_trace3}
                \caption{After applying rule 3 at node $@5$. Replace fact reaches node $@6$.}
                \label{fig:btree_trace3}
        \end{subfigure}%
        ~
        \begin{subfigure}[b]{0.45\textwidth}
                  \includegraphics[width=\textwidth]{btree_trace4}
                  \caption{After applying rule 1 at node $@6$. Value of key 6 has changed to 7.}
                  \label{fig:btree_trace4}
          \end{subfigure}
        \caption{An execution trace for the binary tree dictionary algorithm.}\label{fig:btree_trace}
\end{figure*}

@@

\renewcommand{\arraystretch}{1.2}
\begin{table}[h]
\centering
{\scriptsize
\begin{tabular}{ l l c l }
  Program & $Prog$ & $::=$ & $\Sigma, D$ \\
  Set Of Rules & $\Sigma$ & $::=$ & $\cdot \; | \; \Sigma, R$\\
  Database & $D$ & $::=$ & $\Gamma; \Delta$ \\
  Rule & $R$ & $::=$ & $BE \lolli HE \; | \; \forall_{x}. R$ \\
  Body Expression & $BE$ & $::=$ & $L \; | \; P \; | \; C \; | \; BE, BE \; | \; \exists_{x}. BE \; | \; 1$\\
  Head Expression & $HE$ & $::=$ & $L \; | \; P \; | \; HE, HE \; | \; CE \; | \; AE \; | \; 1$\\

  Linear Fact & $L$ & $::=$ & $l(\hat{x})$\\
  Persistent Fact & $P$ & $::=$ & $\bang p(\hat{x})$\\
  Constraint & $C$ & $::=$ & $c(\hat{x})$ \\

  Comprehension & $CE$ & $::=$ & $\comprehension$ \\
  Aggregate & $AE$ & $::=$ & $\aggregate$ \\
  Aggregate Operation & $A$ & $::=$ & $\mathtt{min} \; | \; \mathtt{max} \; | \; \mathtt{sum} \; | \; \mathtt{count}$ \\

  Sub-Head & $SH$ & $::=$ & $L \; | \; P \; | \; SH, SH \; | \; 1$\\

  Linear Facts & $\Delta$ & $::=$ & $\cdot \; | \; \Delta, l(\hat{t})$ \\
  Persistent Facts & $\Gamma$ & $::=$ & $\cdot \; | \; \Gamma, \bang p(\hat{t})$ \\
\end{tabular}
}
\caption{Abstract syntax of LM.}
\label{tbl:ast}
\end{table}
\renewcommand{\arraystretch}{1.0}

Table~\ref{tbl:ast} shows the abstract syntax for rules in LM.
A LM program $Prog$ consists of a set of derivation rules $\Sigma$ and a database $D$.
Each derivation rule $R$ can be written as $BE \lolli HE$ where $BE$ is the body of a rule and
$HE$ is the head. Rules without bodies are allowed in LM and they are called \textit{axioms}. Rules without heads are also allowed.
The body of a rule, $BE$, may contain linear ($L$) and persistent ($P$) \emph{fact expressions}
and constraints ($C$). Fact expressions are template facts that instantiate variables
(from facts in the database). Variables can be used again in the body for matching and
also in the head when instantiating facts. Constraints are boolean expressions that must
be true in order for the rule to be fired. Constraints use variables from fact expressions and are built using a small functional language that includes mathematical operations, boolean operations, external functions and literal values.
The head of a rule, $HE$, contains linear ($L$) and persistent ($P$) \emph{fact templates} which are uninstantiated facts to derive new facts. The head can also have \emph{comprehensions} ($CE$) and \emph{aggregates} ($AE$). Head
expressions may use the variables instantiated in the body.

\subsubsection{Comprehensions}

Sometimes we need to consume a linear fact and then immediately generate several facts depending on
the contents of the database. To solve this particular need, we created the concept of comprehensions, which are
sub-rules that are applied with all possible combinations of facts from the database. In a comprehension $\comprehension$,
$\widehat{x}$ is a list of variables, $BE$ is the body of the comprehension and $SH$ is the head.
The body $BE$ is used to generate all possible combinations for the head $SH$, according to the facts
in the database.

The following example illustrates a simple program that uses comprehensions:

{\footnotesize
\begin{Verbatim}
!edge(@1, @2).
!edge(@1, @3).
iterate(@1).
iterate(A) -o {B | !edge(A, B) | perform(B)}.
\end{Verbatim}
}

When the rule is fired, we consume \texttt{iterate(@1)} and then generate the comprehension. Here, we iterate through
all the \texttt{edge/2} facts that match \texttt{!edge(@1, B)}, which are: \texttt{!edge(@1, @2)} and \texttt{!edge(@1, @3)}.
For each fact, we derive \texttt{perform(B)}, namely: \texttt{perform(@2)} and \texttt{perform(@3)}.

\subsubsection{Aggregates}

Another useful feature in logic programs is the ability to reduce several facts into a single fact.
In LM we have aggregates ($AE$), a special kind of sub-rule that works very similarly to comprehensions.
In the abstract syntax of the aggregate $\aggregate$, $A$ is the aggregate operation, $\widehat{x}$ is the list of variables
introduced in $BE$, $SH_1$ and $SH_2$ and $y$ is the variable in the body
$BE$ that represents the values to be aggregated using $A$. Like comprehensions,
we use $\widehat{x}$ to try all the combinations of $BE$, but, in addition to deriving $SH_1$ for each combination,
we aggregate the values represented by $y$ and derive $SH_2$ only once using $y$.
As an example, consider the following program:

{\footnotesize
\begin{Verbatim}
price(@1, 3).
price(@1, 4).
price(@1, 5).
count-prices(@1).
count-prices(A) -o
   [sum => P | . | price(A, P) | 1 | total(A, P)].
\end{Verbatim}
}

By applying the rule, we consume \texttt{count-prices(@1)} and
derive the aggregate which consumes all the \texttt{price(@1, P)} facts.
These are summed and \texttt{total(@1,~12)} is derived.  
LM provides several aggregate operations, including the \emph{minimum}, \emph{maximum}, \emph{sum}, and \emph{count}.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\section{The LM Language}
\label{lm_language}

Linear Meld (LM) is a logic programming language that offers a
declarative and structured way to manage mutable state.  A program consists of
a database of facts and a set of derivation rules. The database
includes persistent and linear facts. Persistent facts cannot be
deleted, while linear facts can be asserted and retracted.

The dynamic (or operational) semantics of LM is identical to
Datalog. Initially, we populate the database with the \emph{program's axioms} (initial facts)
and then determine which derivation rules can be applied by
using the current database. Once a rule is applied, new facts can be
derive, which are then added to the database. If a rule uses linear
facts, they are retracted from the database. The program stops when
\emph{quiescence} is achieved, i.e., when rules no longer apply.

Each fact is a predicate on a tuple of \emph{values}, where the type
of the predicate prescribes the types of the arguments.  LM rules are
type-checked using the predicate declarations in the header of the
program. LM has a simple type system that includes types such as
\emph{node}, \emph{int}, \emph{float}, \emph{string},
\emph{bool}. Recursive types such as \emph{list X} and \emph{pair X;
  Y} are also allowed.  Each rule in LM has a defined priority that is
inferred from its position in the source file.  Rules at the beginning
of the file have higher priority.
We consider all the new facts that have been not used yet to create a
set of \emph{candidate rules}.
The set of candidate rules is then applied (by priority)
and updated as new facts are derived.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\begin{figure}[t]
{\footnotesize
\begin{Verbatim}[numbers=right]
type left(node, node).
type right(node, node).
type linear value(node, int, string).
type linear replace(node, int, string).

// set of rules
replace(A, K, New),
value(A, K, Old)
   -o value(A, K, New). // we found our key

replace(A, RKey, RValue),
value(A, Key, Value),
!left(A, B),
RKey < Key
   -o value(A, Key, Value),
      replace(B, RKey, RValue). // go left

replace(A, RKey, RValue),
value(A, Key, Value),
!right(A, B),
RKey > Key
   -o value(A, Key, Value),
      replace(B, RKey, RValue). // go right

// initial configuration
!left(@0, @1).   !right(@0, @2).
!left(@1, @3).   !right(@1, @4). 
!left(@2, @5).   !right(@2, @6).

value(@0, 3, a).   value(@1, 1, b).
value(@2, 5, c).   value(@3, 0, d).
value(@4, 2, e).   value(@5, 4, f).
value(@6, 6, g).

// update key 6 to value x
replace(@0, 6, x).
\end{Verbatim}
}
\caption{LM program for replacing a key's value in a binary tree dictionary}
\label{code:btree_replace}
\end{figure}

We now present in Fig.~\ref{code:btree_replace}, a LM program that
implements the update operation for a binary tree dictionary
represented as key/value pairs. We first declare the predicates (lines
1-4) which represent the facts we are going to use. Predicate
\texttt{left/2} and \texttt{right/2} are persistent while predicates
\texttt{value/3} and \texttt{replace/3} are linear. Predicate
\texttt{value/3} assigns a key/value pair to a tree node and predicate
\texttt{replace/3} represents an update operation that updates the key
in the second argument to the value in the third argument.

The algorithm uses three rules for the three cases of updating a key's
value: the first rule performs the update (lines 6-9); the second rule
recursively picks the left branch for the update operation (lines
11-16); and the third rule picks the right branch (lines 18-23). The
initial axioms are presented in lines 26-33 and they describe the
initial binary tree configuration, including keys and values.
Finally, with the \texttt{replace(@0, 6, x)} axiom instantiated at the
root node $@0$, we intend to change the value of key 6 to value
`$x$'. Note that when writing rules or axioms, persistent facts are
preceded with a \texttt{!}.

\begin{figure*}[t]
\centering
\subfigure[Initial database]
   {\label{fig:btree_trace1}\includegraphics[width=0.44\textwidth]{figures/btree_trace1.pdf}}
\subfigure[After applying rule 3 at node $@0$]
   {\label{fig:btree_trace2}\includegraphics[width=0.44\textwidth]{figures/btree_trace2.pdf}}
\subfigure[After applying rule 3 at node $@2$]
   {\label{fig:btree_trace3}\includegraphics[width=0.44\textwidth]{figures/btree_trace3.pdf}}
\subfigure[After applying rule 1 at node $@6$]
   {\label{fig:btree_trace4}\includegraphics[width=0.44\textwidth]{figures/btree_trace4.pdf}}
\caption{An execution trace for the binary tree dictionary program}
\label{fig:btree_trace}
\end{figure*}

Figure~\ref{fig:btree_trace} illustrates the trace of the
execution. Note that the program database is partitioned by the tree
nodes using the first argument of each fact. In
Fig.~\ref{fig:btree_trace1}, we present the database filled with the
program's axioms. Next, we follow the right branch using rule 3 since
$6 > 3$ (Fig.~\ref{fig:btree_trace2}). We then use the same rule
again in Fig.~\ref{fig:btree_trace3} where we finally reach the key
6. Here, we apply rule 1 and \texttt{value(@6, 6, g)} is updated to
\texttt{value(@6, 6, x)} (Fig.~\ref{fig:btree_trace4}).

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

Table~\ref{tbl:ast} shows the abstract syntax for rules in LM.  An LM
program $Prog$ consists of a set of derivation rules $\Sigma$ and a
database $D$.  Each derivation rule $R$ can be written as $BE \lolli
HE$ where $BE$ is the body of a rule and $HE$ is the head. Rules
without bodies are allowed in LM and they are called
\textit{axioms}. Rules without heads are also allowed.  The body of a
rule, $BE$, may contain linear ($L$) and persistent ($P$) \emph{fact
  expressions} and constraints ($C$). Fact expressions are template
facts that instantiate variables (from facts in the
database). Variables can be used again in the body for matching and
also in the head when instantiating facts. Constraints are boolean
expressions that must be true in order for the rule to be
fired. Constraints use variables from fact expressions and are built
using a small functional language that includes mathematical
operations, boolean operations, external functions and literal values.
The head of a rule, $HE$, contains linear ($L$) and persistent ($P$)
\emph{fact templates} which are uninstantiated facts to derive new
facts. Head expressions may use the variables instantiated in the
body. The head can also have \emph{comprehensions} ($CE$) and
\emph{aggregates} ($AE$).

\begin{table}[ht]
\centering
{\scriptsize
\begin{tabular}{llcl}
Program          & $Prog$ & $::=$ & $\Sigma, D$ \\
Set of Rules     & $\Sigma$ & $::=$ & $\cdot \; | \; \Sigma, R$\\
Database         & $D$ & $::=$ & $\Gamma; \Delta$ \\
Rule             & $R$ & $::=$ & $BE \lolli HE \; | \; \forall_{x}. R$ \\
Body Expression  & $BE$ & $::=$ & $L \; | \; P \; | \; C \; | \; BE, BE \; | \; \exists_{x}. BE \; | \; 1$\\
Head Expression  & $HE$ & $::=$ & $L \; | \; P \; | \; HE, HE \; | \; CE \; | \; AE \; | \; 1$\\
Linear Fact      & $L$ & $::=$ & $l(\hat{x})$\\
Persistent Fact  & $P$ & $::=$ & $\bang p(\hat{x})$\\
Constraint       & $C$ & $::=$ & $c(\hat{x})$ \\
Comprehension    & $CE$ & $::=$ & $\comprehension$ \\
Aggregate        & $AE$ & $::=$ & $\aggregate$ \\
Operations       & $A$ & $::=$ & $\mathtt{min} \; | \; \mathtt{max} \; | \; \mathtt{sum} \; | \; \mathtt{count}$ \\
Sub-Head         & $SH$ & $::=$ & $L \; | \; P \; | \; SH, SH \; | \; 1$\\
Linear Facts     & $\Delta$ & $::=$ & $\cdot \; | \; \Delta, l(\hat{t})$ \\
Persistent Facts & $\Gamma$ & $::=$ & $\cdot \; | \; \Gamma, \bang p(\hat{t})$ \\
\end{tabular}
}
\caption{Abstract syntax of LM}
\label{tbl:ast}
\end{table}

We created the concept of comprehensions to be used when the
consumption of a linear fact should generate a set of facts
accordingly to the current contents of the database. In a
comprehension $\comprehension$, $\widehat{x}$ is a list of variables,
$BE$ is the body of the comprehension and $SH$ is the head. The body
$BE$ is used to generate all possible combinations for the head $SH$,
according to the facts in the database. The following program shows a
simple example that uses comprehensions:

{\footnotesize
\begin{Verbatim}
!edge(@1, @2).   !edge(@1, @3).
iterate(@1).

iterate(A)
   -o {B | !edge(A, B) | perform(B)}.
\end{Verbatim}
}

When the rule is fired, we consume \texttt{iterate(@1)} and then
generate the comprehension. Here, we iterate through all the
\texttt{edge/2} facts that match \texttt{!edge(@1, B)}, which are
\texttt{!edge(@1, @2)} and \texttt{!edge(@1, @3)}. For each fact, we
then derive \texttt{perform(B)}, namely \texttt{perform(@2)} and
\texttt{perform(@3)} in this example.

Another useful feature is the ability to reduce several facts into a
single fact. For that, we have aggregates, a special kind of sub-rule
that works very similarly to comprehensions. In a aggregate
$\aggregate$, $A$ is the aggregate operation, $\widehat{x}$ is the
list of variables introduced in $BE$, $SH_1$ and $SH_2$ and $y$ is the
variable in the body $BE$ that represents the values to be aggregated
using $A$. Like comprehensions, we use $\widehat{x}$ to try all the
combinations of $BE$, but, in addition to deriving $SH_1$ for each
combination, we aggregate the values represented by $y$ and derive
$SH_2$ only once using $y$. As an example, consider the following
program:

{\footnotesize
\begin{Verbatim}
price(@1, 3).   price(@1, 4).   price(@1, 5).
count-prices(@1).

count-prices(A)
   -o [sum => P | . | price(A, P) | 1 | total(A, P)].
\end{Verbatim}
}

By applying the rule, we consume \texttt{count-prices(@1)} and derive
the aggregate which consumes all the \texttt{price(@1, P)} linear
facts. These are summed up and \texttt{total(@1,~12)} is derived. LM
provides several aggregate operations, including the \emph{minimum},
\emph{maximum}, \emph{sum} and \emph{count}.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2014-07-25 paper.tex
Update.

% This is LLNCS.DEM the demonstration file of
% the LaTeX macro package from Springer-Verlag
% for Lecture Notes in Computer Science,
% version 2.4 for LaTeX2e as of 16. April 2010
%
\documentclass{acm_proc_article-sp}
%
\usepackage{makeidx}  % allows for indexgeneration
\usepackage{fancyvrb}
\usepackage{amsmath}
\usepackage{graphicx}
\usepackage{caption}
\usepackage{subcaption}
\usepackage{wrapfig}
\usepackage{nth}
\usepackage[backref,pageanchor=true,plainpages=false, pdfpagelabels, bookmarks,bookmarksnumbered]{hyperref}
\input{fp-macros}

\begin{document}
\newcommand{\cmu}{\ensuremath{^\dag}}
\newcommand{\fcup}{\ensuremath{^\ddag}}
  \makeatletter
    \let\@copyrightspace\relax
      \makeatother
\numberofauthors{3} %  in this sample file, there are a *total*

\title{Design and implementation of a multithreaded virtual machine for executing linear logic programs}

\author{
\alignauthor
   Flavio Cruz \\
\affaddr{Carnegie Mellon University, Pittsburgh, PA 15213, USA}\\
\email{fmfernan@cs.cmu.edu} \\
\alignauthor Ricardo Rocha \\
\affaddr{CRACS \& INESC TEC, Faculty of Sciences, University Of Porto} \\
\affaddr{Rua do Campo Alegre, 1021/1055, 4169-007 Porto, Portugal} \\
\email{ricroc@dcc.fc.up.pt} \\
\alignauthor Seth Copen Goldstein \\
\affaddr{Carnegie Mellon University, Pittsburgh, PA 15213, USA}\\
\email{seth@cs.cmu.edu} \\
}

%\institute{Carnegie Mellon University, Pittsburgh, PA 15213, USA\\
%\email{\{fmfernan, seth\}@cs.cmu.edu}
%\and
%CRACS \& INESC TEC, Faculty of Sciences, University Of Porto\\
%Rua do Campo Alegre, 1021/1055, 4169-007 Porto, Portugal\\
%\email{\{flavioc, ricroc\}@dcc.fc.up.pt}}

\maketitle              % typeset the title of the contribution

\begin{abstract}
\input{abstract}
\keywords{linear logic programming, virtual machine, implementation}
\end{abstract}

\section{Introduction}
\input{introduction}

\section{The LM Language}\label{lm_language}
\input{language}

\section{The Virtual Machine}\label{virtual_machine}
\input{virtual_machine}

\section{Experimental Results}\label{results}
\input{preliminary_results}

\section{Related Work}\label{related_work}
\input{related_work}

\section{Conclusions}
\input{conclusions}

>>>>

\documentclass{acm_proc_article-sp}
\usepackage[utf8]{inputenc}
\usepackage{epsfig}
\usepackage{multicol}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{subfigure}
\usepackage{fancyvrb}
\usepackage{url}
\usepackage{balance}
\usepackage{listings}
\usepackage{multicol}
\lstset{language=C, basicstyle=\ttfamily, breaklines=true, frame=lines}
\input{fp-macros}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\title{Design and Implementation of a Multithreaded Virtual Machine for Executing Linear Logic Programs}

\numberofauthors{3}
\author{
  \alignauthor Flavio Cruz \\
    \affaddr{Carnegie Mellon University}
    \affaddr{Pittsburgh, PA 15213, USA}\\
    \email{fmfernan@cs.cmu.edu} \\
  \alignauthor Ricardo Rocha \\
    \affaddr{CRACS \& INESC TEC}\\
    \affaddr{University of Porto}\\
    \affaddr{Rua Campo Alegre 1021/1055}\\
    \affaddr{4169-007 Porto, Portugal}\\
    \email{ricroc@dcc.fc.up.pt}
  \alignauthor Seth Copen Goldstein \\
    \affaddr{Carnegie Mellon University}
    \affaddr{Pittsburgh, PA 15213, USA}\\
    \email{seth@cs.cmu.edu} \\
}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\maketitle

\input{abstract}
\input{introduction}
\input{language}
\input{virtual_machine}
\input{experiments}
\input{related_work}
\input{conclusions}

\balance

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

### 2014-07-25 related_work.tex
Update.

@@

Prolog, the most prominent logic programming language, has a rich history of virtual machine research centered
around the Warren Abstract Machine WAM~\cite{AICPub641:1983}. The WAM offers special purpose instructions, including
unification instructions for different kinds of data and control flow instructions to implement backtracking.
The WAM is fully sequential and uses four memory areas: heap, stack, trail and the push down list.
Much research has been done to improve the speed and efficiency of the original WAM design~\cite{Costa07demand-drivenindexing,167005,Turk-Logic:1986fk}.

Prolog is naturally parallel because several clauses for the same goal (AND-parallelism) or all goals in a clause (OR-parallelism) can be tried
in parallel. Different abstract machines for AND-parallelism has been developed on top of the WAM~\cite{Hermenegildo:1986:AMB:913061,Lin:1988:AEL:900478}.
For OR-parallelism we have several models such as: the SRI model~\cite{Warren:1987:OEM:67683.67699},
the Argonne model~\cite{ButlerDLOOS88}, the MUSE model~\cite{Ali:1990fk} and the BC machine~\cite{Ali88}. While all those models have been
developed using the WAM, there are some parallel machines totally different from the WAM such as the PPAM~\cite{Kacsuk:1990:EMP:533578},
which is based on a data-flow model.

>>>>

\section{Related Work}
\label{related_work}

@@

Prolog, the most prominent logic programming language, has a rich
history of virtual machine research centered around the Warren
Abstract Machine (WAM)~\cite{AICPub641:1983}. 
%% The WAM offers special purpose instructions, including unification
%% instructions for different kinds of data and control flow instructions
%% to implement backtracking.  The WAM is fully sequential and uses four
%% memory areas: heap, stack, trail and the push down list.  Much
%% research has been done to improve the speed and efficiency of the
%% original WAM
%% design~\cite{Costa07demand-drivenindexing,167005,Turk-Logic:1986fk}.
Prolog is naturally parallel because several clauses for the same goal
(AND-parallelism) or all goals in a clause (OR-parallelism) can be
tried in parallel. Different abstract machines for AND-parallelism has
been developed on top of the
WAM~\cite{Hermenegildo:1986:AMB:913061,Lin:1988:AEL:900478}.  For
OR-parallelism we have several models such as: the SRI
model~\cite{Warren:1987:OEM:67683.67699}, the Argonne
model~\cite{ButlerDLOOS88}, the MUSE model~\cite{Ali:1990fk} and the
BC machine~\cite{Ali88}. While all those models have been developed
using the WAM, there are also some parallel machines totally different
from the WAM, such as the PPAM~\cite{Kacsuk:1990:EMP:533578}, which is
based on a data-flow model.
